---
title: "Week 1 - Donny Lofland"
output:
  word_document: default
  html_notebook: default
  pdf_document: default
  html_document:
    df_print: paged
---
```{r}
# SOURCE: https://rpubs.com/djlofland/math_workshop_1
library(Deriv)
library(mosaic)
library(mosaicCalc)
library(rSymPy)
```

Find the derivatives with the respect to x of the following.

1. F(x|x≥0)=$1−e^{(−\lambda x)}$
```{r}
myf1=function(x){1-exp(-lambda*x)}
print(Deriv(myf1))
```
***
2. $F(x|b>a)=\frac{(x−a)}{(b−a)}$
```{r}
myf2 = function(x){(x-a)/(b-a)}
print(Deriv(myf2))
```
***
3. $F(x|a<x≤c≤b)=\frac{(x−a)^2}{(b−a)(c−a)}$
```{r}
myf3 = function(x){(x-a)^2/((b-a)(c-a))}
print(Deriv(myf3))
```
***
4. $F(x|a≤c<x<b)=1−\frac{(b−x)^2}{(b−a)(c−a)}$
```{r}
myf4 = function(x)1-(b-x)^2/((b-a)(c-a))
print(Deriv(myf4))
```

***
Solve the following definite and indefinite integrals

5. $\int_0^{10} 3x^3dx$
```{r}
myf5=function(x)3*x^3
integrate(Vectorize(myf5),0,10)
```
***
6. $\int_0^x x\lambda e^{−\lambda x}dx$

Solution:

Substitution
$u=-\lambda x$; 
$du=-\lambda dx$; 
$dx=\frac{du}{-\lambda}$; 

$\int -u e^u \frac{du}{-\lambda}$;

$\frac{1}{\lambda}\int u e^u du$

solve $\int u e^u du$ by Parts:
$s=u; ds=du$; 
$dv=e^udu; v=e^u+C$; 

formula: $sv-\int vds$

$=u e^u - \int e^u du$

$=u e^u - e^u$

$= (u-1)e^u$

plugging back in: $\frac{1}{\lambda}(u-1)e^u$ 

replacing $u=-\lambda x$ back in: $\frac{1}{\lambda}\int u e^u du$

$=\frac{1}{\lambda}(-\lambda x - 1)e^{-\lambda x}$

$= -\frac{(\lambda x + 1) e^{-\lambda x}}{\lambda}$

lastly we need to integrate over $0\to x$:

$-\frac{(\lambda x + 1) e^{-\lambda x}}{\lambda} ]_0^x$
$=-\frac{(\lambda x + 1) e^{-\lambda x}}{\lambda} - \frac{1}{\lambda}$
$=-\frac{(\lambda x + 1) e^{-\lambda x}+1}{\lambda}$


```{r}
library(rSymPy)
sympy("x = Symbol('x')")
sympy("l = Symbol('lambda')")

sympy("integrate(x*l*exp(-l*x),(x,0,x))")
```
Answer from sympy: $\frac{-\lambda e^{-\lambda x}}{\lambda^2}-\frac{x e^{-\lambda x}}{\lambda}+\frac{1}{\lambda}$

I'm not sure how to get r to calculate this or it was intended that we to this by hand.  I made my best stab at doing by hand and code in latex for the notebook.

***
7. $\int_0^{0.5} \frac{1}{b−a}dx$

Solution:
$\frac{1}{b-a} x ]_0^x = \frac{x}{b-a} - 0 = \frac{x}{b-a}$

```{r}
library(rSymPy)
sympy("x = Symbol('x')")
sympy("a = Symbol('alpha')")
sympy("B = Symbol('beta')")

sympy("integrate(1/(B-a),(x,0,x))")
```
$\frac{x}{b-a}$

I'm not sure how to get r to calculate this or it was intended that we to this by hand.  I'm a python guy so brought over sympy to help.  I made my best stab at doing by hand and code in latex for the notebook.

***
8. $\int_0^x x \frac{1}{\Gamma(\alpha)\beta^\alpha} x^{\alpha -1} e^{-\beta x} dx$

Solution:
```{r}
library(rSymPy)
sympy("x = Symbol('x')")
sympy("a = Symbol('alpha')")
sympy("B = Symbol('beta')")
sympy("G = Symbol('Gamma')")

sympy("integrate(x * (G*a*B**a)**(-1) * x**(a) * exp(-B*x), (x,0,x))")
```
sympy solution: $\frac{x \beta^{-\alpha}x^{\alpha-1}e^{-\beta x}}{\Gamma \alpha}]_0^x$

I really have no idea how to tackle this or was was hinted by the gamma function.  I'm hoping we can cover this at some point so I can understand exactly how to approach the problem.  Did you want us to hand solve this, use r or some combination?  I had assumed this was a r task since we were using R to solve the early problems.  Anyways, for problem 7 & 8, I could certainly use some insights.  Thanks! Donny


Hint: the last part of the equation is beginning with the gamma function is a Gamma probability distribution function. Try rearranging the terms to integrate another Gamma distribution out of the integral, as pdfs must integrate to 1.

***

With the following matrix,
$X = \begin{bmatrix}1 & 2 & 3 \\ 3 & 3 & 1 \\ 4 & 6 & 8\end{bmatrix}$
```{r}
myMatrix <- matrix(c(1,3,4,2,3,6,3,1,8),3,3)
myMatrix
```
***
9. Invert it using Gaussian row reduction.
```{r}
library(matlib)
gaussianElimination(myMatrix, numeric(3))
gaussianElimination(myMatrix, diag(3))
inv(myMatrix)
```
***
10. Find the determinant.
```{r}
det(myMatrix)
```
***
11. Conduct LU decomposition
```{r}
lum <- lu(myMatrix)
elu <- expand(lum)

(L <- elu$L)
(U <- elu$U)
(P <- elu$P)

L %*% U
```
***
12. Multiply the matrix by it’s inverse.
```{r}
inv <- round(solve(myMatrix),2)
myMatrix%*%inv
```

