---
title: 'R Notebook: Projeto de Economia Aplicada - academic year 2021/2022'
author: 'Carla Sá & Miguel Portela - v.miguel'
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

> EDA WITH R

## SET THE ENV VARIABLES FOR PLOTLY.COM

```{r}
rm(list = ls())
gc()

setwd("/Users/miguelportela/Documents/GitHub/R_Training/EDA/_theomitsa_example")
Sys.setenv("plotly_username"="theomitsa")
Sys.setenv("plotly_api_key"="Ojps7ld5FByuyrpKV2zC")

library(tidyverse)

```


## READ THE INPUT FILE

```{r}
sal<-read.csv("Salaries2.csv",header=TRUE)

```

## QUICK DATAFRAME CHECK

### Dimensions of the dataset

```{r}
dim(sal)

```

### Number and Types of Variables

```{r}
str(sal)

```

### Duplicates
### How many are the unique values

```{r}
uu<-unique(sal)
dim(uu)
dim(sal)

```


### Which Are the Duplicates

```{r}
library(janitor)
dupj<-get_dupes(sal)
dupj

```

### Missing Data

### Print the no. of complete rows and total no.of missing cases

```{r}
library(DataExplorer)
introduce(sal)

```

### For each column print the number of missing cases

```{r}
apply(apply(sal,2,is.na),2,sum)

```

### Print the cases(rows) with the missing values

```{r}
sal[!complete.cases(sal),]

```

### UNIVARIATE STATISTICS

```{r}
summary(sal)

```

### psych package

```{r}
library(psych)
psych::describe(sal)

```

### dataMaid package

```{r}
library(dataMaid)
summarize(sal)
makeDataReport(sal, replace=TRUE,output="gg.html")

```

### Additional univariate description functions

```{r}
library(summarytools)
dfSummary(sal)

```

```{r}
library(inspectdf)
inspect_num(sal)
inspect_cat(sal)

```

```{r}
library(prettyR)
prettyR::describe(sal)

```

```{r}
library(Hmisc)
Hmisc::describe(sal)

```

```{r}
library(funModeling)
profiling_num(sal)

```

```{r}
library(skimr)
skim(sal)

```

## Univariate Group Statistics
### Use tapply() to group by a factor

```{r}
tapply(sal$salary,sal$sex,mean,na.rm=TRUE)

```

> Within groups, print detailed statistical information of the var under examination.

```{r}
library(psych)
describeBy(sal$salary,group=sal$sex)

```

### Use dplyr functions

```{r}
newgroup<-sal%>%filter(discipline=='A')%>%
  group_by(sex,rank)%>%summarise(Avgsal=mean(salary))

newgroup

```

### Plots of Univariate Statistics (boxplot,histogram)

```{r}
ggplot(sal,aes(sex,salary))+geom_boxplot()
ggplot(sal,aes(x=rank, fill=sex))+geom_bar()

```

### BIVARIATE STATISTICS

### Interactive Scatter Plot

```{r}
library(plotly)
gg1<-ggplot(na.omit(sal), aes(salary,yrs.since.phd, col=sex))+geom_point()+geom_smooth(method=lm)
p1<-ggplotly(gg1)
api_create(p1,filename="plot of salary vs yrs.since.phd")

```

### Interactive 3D plot

```{r}
library(plotly)

#fig<-plot_ly(sal, x=~salary,y=~yrs.service,z=~yrs.since.phd,color=~sex,color_discrete_map={"female":"red"})%>% add_markers()

#fig
#api_create(fig,filename="3D plot")

```



## Correlations

### Correlations displayed in a table with package stats

```{r}
sal1<-select_if(sal, is.numeric)
cor(na.omit(sal1))
library(stats)
cor.test(sal1$salary,sal1$yrs.service)

```

## Correlation plots

### Corrplot package

```{r}
library(corrplot)
cc<-cor(na.omit(sal1))
corrplot(cc,method="pie")

```

### PerformanceAnalytics package

```{r}
library(PerformanceAnalytics)
chart.Correlation(sal1, histogram=TRUE, pch=19)

```

### ggcorrplot package

```{r}
library(ggcorrplot)
ggcorrplot(cor(na.omit(sal1)), p.mat = cor_pmat(na.omit(sal1)), hc.order=TRUE, type='lower')
cor_pmat(na.omit(sal1))

```

## ALL INCLUSIVE CORRELATION

### DataExplorer package

```{r}
library(DataExplorer)
plot_correlation(sal)

```

### psych package

```{r}
library(psych)
pairs.panels(sal,bg=c("yellow","blue")[sal$sex],pch=21, cex.cor=6, scale=TRUE)

```

### GGally package

```{r}
library(GGally)
ggpairs(sal)

```

### Test of Independence

```{r}
table(sal$rank)
table(sal$rank,sal$sex)
library(gmodels)
t2<-CrossTable(sal$rank,sal$sex,chisq=TRUE,fisher=TRUE)
t2

```


### T-test of Significant Diferences

```{r echo=FALSE}
filtered<-filter(sal, sex=='Male')
c1<-(filtered[6])
filtered2<-filter(sal, sex=='Female')
c2<-(filtered2[6])
t.test(c1,c2)

```

### T-test with rank as the fixed effect

```{r echo=FALSE}
filtered<-filter(sal, sex=='Male',rank=='Prof')
c1<-(filtered[6])
filtered2<-filter(sal, sex=='Female', rank=='Prof')
c2<-(filtered2[6])
t.test(c1,c2)

```

## OUTLIERS

### Univariate Outlier Detection

### Interactive histogramM

```{r echo=FALSE}
library(plotly)
p1<-ggplot(sal,aes(sex,salary))+geom_boxplot()
p1
pp<-ggplotly(p1)
api_create(pp, filename = "boxplot-salary-sex")

```

### Package extremevalues

```{r echo=FALSE}
library(extremevalues)
L<-getOutliers(sal$salary)
outlierPlot(sal$salary,L)

```

### Package OutlierDetection

```{r eval=FALSE, include=FALSE}
# library(outliers)
#library(OutlierDetection)

#UnivariateOutlierDetection(sal$salary)

```

### Multivariate Outlier Detection

### Package OutlierDetection

```{r echo=FALSE}
#library(OutlierDetection)
sal1<-select(sal,salary,yrs.service)
saldata<-na.omit(sal1)
salr <- saldata[,c("yrs.service", "salary")]
#OutlierDetection(salr)

```

### Package Chemometrics

```{r echo=FALSE}
library(chemometrics)
gg<-Moutlier(saldata)
mahal_dist<-gg$md
names(mahal_dist) <- 1:nrow(saldata)
ff<-sort(mahal_dist, decreasing = TRUE)
ff

```

### Package Mvoutlier

```{r echo=FALSE}
library(mvoutlier)
par("mar")
par(mar=c(1,1,1,1))
out <- chisq.plot(saldata)
out$outliers

```


# AUTOMATIC

## DataExplorer

```{r echo=FALSE, include=FALSE}
library(DataExplorer)
create_report(sal,y="salary")

```

## AEDA

```{r echo=FALSE,include=FALSE}
#library(efficient)
library(AEDA)
fastReport(data=sal)
rmarkdown::render("MainReport.rmd")

```

## ExPanDaR

```{r eval=FALSE, include=FALSE}
library(ExPanDaR)
ExPanD(sal)

```

## RtutoR

```{r eval=FALSE, include=FALSE}
library(RtutoR)
gg=generate_exploratory_analysis_ppt(sal,target_var="salary",output_file_name="Rtutorreport.pptx")

```

