---
title: "Applied Data Analysis School"
author: "Miguel Portela"
date: "October 27 and 29, 2020"
institute: "Universidade do Minho"
output:
  beamer_presentation: default
  slidy_presentation: default
  ioslides_presentation: default
  pdf_document: default
subtitle: LITERATE PROGRAMMING IN R MARKDOWN
theme: Copenhagen
header-includes:
- \usepackage{hyperref}
- \hypersetup{ colorlinks=true, linkcolor=blue, urlcolor=blue, }
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE,results='hide',message=FALSE}
rm(list = ls())
# setwd("/Users/miguelportela/Documents/GitHub/R_Training")
setwd("C:/Users/mangelo.EEG/Documents/GitHub/R_Training")
# install.packages("rmarkdown")
library(tidyverse)
library(ggplot2)
library(haven)
library(rmarkdown)

# data()
attach(mtcars)
nlswork <- read_dta("data/nlswork.dta")
summary(nlswork)
attach(nlswork)
summary(ln_wage)
```

# Setup and references

1. [Pandoc](https://pandoc.org/installing.html)

2. Latex

  - You can install a lightweight Latex flavor: [TinyTex](https://yihui.name/tinytex/)

```{r eval=FALSE}
  install.packages('tinytex')
  tinytex::install_tinytex()  
```

  - or a complete distribution: [MikTex](https://miktex.org/download) or [MacTex](https://tug.org/mactex/)

3. [Sumatra PDF](https://www.sumatrapdfreader.org/download-free-pdf-viewer.html) or [Skim](https://skim-app.sourceforge.io/)

4. Markdown references

  - [Markdown Guide](https://www.markdownguide.org/)

  - [R Markdown - Get Started](https://rmarkdown.rstudio.com/lesson-1.html)

  - [R Rarkdown: The definitive guide](https://bookdown.org/yihui/rmarkdown/)

# By the end of the day

you can write a report

- Claxton, G., Rae, M., Long, M., Damico, A., Foster, G. and Whitmore, H., 2017. [2017 Employer Health Benefits Survey](http://files.kff.org/attachment/Report-Employer-Health-Benefits-Annual-Survey-2017). _Kaiser Family Foundation and Health Research & Educational Trust_.

a paper

- Hartgerink, C. H. J., Wicherts, J. M., & van Assen, M. A. L. M. (2017). [Too Good to be False: Nonsignificant Results Revisited](https://www.collabra.org/articles/10.1525/collabra.71/galley/82/download/). _Collabra: Psychology_, 3(1), 9.

edit a book

- Xie, Y., Allaire, J.J. and Grolemund, G., 2018. [_R Markdown: The Definitive Guide_](https://bookdown.org/yihui/rmarkdown/). CRC Press.

or build your website

- [Rob J Hyndman’s website](https://robjhyndman.com/)

# Introduction

- What is [**Literate programming**?](https://en.wikipedia.org/wiki/Literate_programming)

> Literate programming refers to melding a descriptive narrative and computer code into a single document, from which both
> human-friendly documentation and computer readable files can be created

- Your work should be transparent, easy to update, easy to maintain, and easy to replicate

- Data analysis requires reproducibility of results

- Useful for teaching

# Preliminaries: Pandoc

- [**Pandoc**](https://pandoc.org/) is a command line tool to convert across different document formats. Try it online [**here**](https://pandoc.org/try/)

- It supports many formats such as Markdown, HTML, docx, pdf, $LaTeX$, etc

- It is simple to use: to convert the file *example1.md* to

  - HTML
  ```
  pandoc -s example1.md -o example1.html
  ```
  - PDF
  ```
  pandoc -s example1.md -o example1.pdf
  ```
  - docx
  ```
  pandoc -s example1.md -o example1.pdf
  ```

- Note: `-s` stands for source and `-o` for output

# Preliminaries: Markdown

- [**Markdown**](https://en.wikipedia.org/wiki/Markdown) is a lightweight markup language with plain text formatting syntax that was invented by John Gruber

- There are several "flavors" of Markdown

- Markdown files can be created in any text editor

- and can be easily converted to other formats

- They are text files usually with the extension ".md"

- There are many editors specific for Markdown (eg: [**Typora**](https://typora.io/), [**Mou**](http://25.io/mou/), [**Draft**](https://draftin.com/), etc) but you can use general editors (eg: [**Atom**](https://atom.io/), [**Sublime Text**](https://www.sublimetext.com/) or [**RStudio**](https://rstudio.com/))

- Used in popular sites such as [**GitHub**](https://github.com/), [**Reddit**](https://www.reddit.com/),[**Stack Exchange**](https://stackexchange.com/), etc

# Markdown components

1. **Metadata**: written between a pair of three dashes **`---`**

2. **Text**: regular text and math elements, plus tables, images and figures

3. **Code**:
  
  - _inline R code_: starts with `` `r `` and ends with a backtick `` ` ``
  
  - _code chunk_: starts with three backticks like ```` ```{r} ```` where `r` indicates the language name and ends with three backticks ```` ``` ````

# Markdown syntax: quick introduction

- Creating Headers

```
This is header 1
================
```

or

```
Another header
--------------
```

- or headers with different levels

```
# This is Header 1
...some text...
## And now Header 2
...more text...
### Finally Header 3
... and a little more text.
```

# Markdown syntax (continued)

- Emphasizing text

```
this is *Italic* but _this_ also works
```

```
this will write in **bold** but you can also __do this__
```

```
It is also possible to ~~Strikeout~~ text
```

- Creating unordered lists (use "*" "+" or "-"")
```
* Point 1
+ Point 2
```

# Markdown syntax (continued)

- You can create structured lists

```
- Point 1
  - Point 12
- Point 2
  - Point 21
```

- insert links to text. Simply write
```
[text](url)
```
- insert images
```
![text](file "Description")
```

- You can add tables

- and even references

# Markdown syntax (continued)

- Markdown documents are highly flexible because they can incorporate other languages

- An example is [$LaTeX$](https://en.wikipedia.org/wiki/LaTeX). To add $LaTeX$ code you need to enclose it in "$"
If you write
```
$a^2+b^2=c^2$
```
you will see $a^2+b^2=c^2$

- Markdown also accepts raw HTML code

- You can incorporate code from other languages (eg: Python, Java, R, Stata, etc)

- This is typically done in a "code fence": lines with three or more backticks or tildes inserted before and after

# Markdown syntax (continued)

- You can find a quick guide for syntax [**here**](https://www.markdownguide.org/cheat-sheet/) or more detailed information [**here**](https://daringfireball.net/projects/markdown/syntax)

- To get an idea how Markdown works you can use an online editor such as [**Dillinger**](https://dillinger.io/) or [**Markdown here**](https://markdown-here.com/livedemo.html)

- and if you need to create tables you can use [**this**](http://www.tablesgenerator.com/markdown_tables) great online tool

- You can use the [Cheat sheets](https://www.rstudio.com/resources/cheatsheets/) - also available via the Help menu in RStudio

- For the exercises explore the random text generator [Lorem ipsum](https://loremipsum.io/generator/)


# Metadata

- You can add a title, author and date as metadata. Simply start the document as

```
% Fill in title
% Fill in Name
% 23 January 2019
```

- Or you can use the [**YAML**](https://en.wikipedia.org/wiki/YAML) format. See [**Pandoc User's Guide**](https://pandoc.org/MANUAL.pdf) for more info. To add title, author and date place at the top of the document

```
---
title: Fill in title
author: Fill in name
date: 23 January 2019
---
```

# Definig the output formats

- `pdf_document`

- `html_document`

- `word_document`

- `beamer_presentation`

- `ioslides_presentation`

- `powerpoint_presentation`

```
---
title: Fill in title
author: Fill in name
date: 23 January 2019
output:
  pdf_document:
    toc: true
---
```

# Adding citations to the document

- it is possible to add citations from a [**bibtex**](http://www.bibtex.org/) file

- place the "bib" file (say "references.bib") in the working folder

- modify the YAML by adding a line pointing to the bib file

```
bibliography: references.bib
```

- At the bottom of the document add a line

`## References`

- use the syntax [@key] to identify the reference

# Advanced use: customizing outputs

- The default style for citations is the Chicago Manual of Style author-date format.

- But you can use any style available in Citation Style Language (CSL) in the [Zotero Style Repository](https://www.zotero.org/styles).

- To change style download the "csl" file and add a reference to it in the YAML block. for example

```{r eval=FALSE}
---
title: Title of my presentation
author: My name
date: 23 January 2019
bibliography: references.bib
csl: thisstyle.csl
---
```

# Further information

- [LITERATE PROGRAMMING IN R MARKDOWN: GitHub repository](https://github.com/reisportela/R_Training)

- [Project available in RStudio Cloud](https://rstudio.cloud/spaces/100541/join?access_code=znpd9DEjghnfbdj%2FvHVjuOTQA0gjkVIDI1xbQXIC)
