---
title: 'Notes on EDA'
author: 'Miguel Portela - V.01'
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

> EDA WITH R

## SET THE ENV VARIABLES FOR PLOTLY.COM

```{r, setup, echo=FALSE, include=FALSE}
  gc()
  rm(list = ls())
  knitr::opts_knit$set(root.dir = '/Users/miguelportela/Documents/GitHub/R_Training/EDA/')
  library(tidyverse)
  library(haven)
  library(readxl)

Sys.setenv("plotly_username"="reisportela")
Sys.setenv("plotly_api_key"="3bwfj4aFVl7VcTphTddJ")

```


## READ THE INPUT FILE

```{r}
sal<-read.csv("_theomitsa_example/Salaries2.csv",header=TRUE)

df_base <- read_excel("nlswork.xlsx")
df <- as.data.frame(df_base)
```

## QUICK DATAFRAME CHECK

### Dimensions of the dataset

```{r}
dim(sal)
dim(df)

```

### Number and Types of Variables

```{r}
str(sal)
str(df)

```

### Duplicates
### How many are the unique values

```{r}
uu<-unique(sal)
dim(uu)

uu<-unique(df)
dim(uu)
```


### Which Are the Duplicates

```{r}
library(janitor)
dupj<-get_dupes(sal)
dupj

dupj<-get_dupes(df)
dupj
```

### Missing Data

### Print the no. of complete rows and total no.of missing cases

```{r}
library(DataExplorer)
introduce(sal)
introduce(df)
```

### For each column print the number of missing cases

```{r}
apply(apply(sal,2,is.na),2,sum)
apply(apply(df,2,is.na),2,sum)

```

### Print the cases(rows) with the missing values

```{r}
sal[!complete.cases(sal),]
df[!complete.cases(df),]

```

### UNIVARIATE STATISTICS

```{r}
summary(sal)
summary(df)

```

### psych package

```{r}
library(psych)
psych::describe(sal)
psych::describe(df)

```

### dataMaid package

```{r}
library(dataMaid)
summarize(sal)
makeDataReport(sal, replace=TRUE,output="dM_sal.html")

summarize(df)
makeDataReport(df, replace=TRUE,output="dM_nlswork.html")

```

### Additional univariate description functions

```{r}
library(summarytools)
dfSummary(sal)

dfSummary(df)
```

```{r}
library(inspectdf)
inspect_num(sal)
inspect_cat(sal)

inspect_num(df)
inspect_cat(df)
```

```{r}
library(prettyR)
prettyR::describe(sal)

prettyR::describe(df)
```

```{r}
library(Hmisc)
Hmisc::describe(sal)

Hmisc::describe(df)
```

```{r}
library(funModeling)
profiling_num(sal)

profiling_num(df)
```

```{r}
library(skimr)
skim(sal)

skim(df)
```

## Univariate Group Statistics
### Use tapply() to group by a factor

```{r}
tapply(sal$salary,sal$sex,mean,na.rm=TRUE)

tapply(df$ln_wage,df$union,mean,na.rm=TRUE)
```

> Within groups, print detailed statistical information of the var under examination.

```{r}
library(psych)
describeBy(sal$salary,group=sal$sex)

describeBy(df$ln_wage,group=df$union)
```

### Use dplyr functions

```{r}
newgroup<-sal%>%filter(discipline=='A')%>%
  group_by(sex,rank)%>%summarise(Avgsal=mean(salary))

  newgroup

newgroup<-df%>%filter(south==1)%>%
  group_by(union,collgrad)%>%summarise(Avgsal=mean(grade))

  newgroup

```

### Plots of Univariate Statistics (boxplot,histogram)

```{r}
ggplot(sal,aes(sex,salary))+geom_boxplot()
ggplot(sal,aes(x=rank, fill=sex))+geom_bar()

ggplot(df,aes(collgrad,age))+geom_boxplot()
ggplot(df,aes(x=collgrad, fill=union))+geom_bar()
```

### BIVARIATE STATISTICS

### Interactive Scatter Plot

```{r}
library(plotly)
## CHECK 'API' here: https://plotly.com/r/getting-started-with-chart-studio/

gg1<-ggplot(na.omit(sal), aes(salary,yrs.since.phd, col=sex))+geom_point()+geom_smooth(method=lm)
p1<-ggplotly(gg1)
api_create(p1,filename="plot of salary vs yrs.since.phd")

# gg2<-ggplot(na.omit(df), aes(ln_wage,hours, col=south))+geom_point()+geom_smooth(method=lm)
# p2<-ggplotly(gg2)
# api_create(p2,filename="plot of ln_wage vs hours")

```

### Interactive 3D plot

```{r}
library(plotly)

#fig<-plot_ly(sal, x=~salary,y=~yrs.service,z=~yrs.since.phd,color=~sex,color_discrete_map={"female":"red"})%>% add_markers()

#fig
#api_create(fig,filename="3D plot")

```



## Correlations

### Correlations displayed in a table with package stats

```{r}
sal1<-select_if(sal, is.numeric)
cor(na.omit(sal1))

library(stats)
cor.test(sal1$salary,sal1$yrs.service)

df1<-select_if(df, is.numeric)
cor(na.omit(df1))

cor.test(df1$ln_wage,df1$age)
```

## Correlation plots

### Corrplot package

```{r}
library(corrplot)
cc<-cor(na.omit(sal1))
corrplot(cc,method="pie")

ss<-cor(na.omit(df1))
corrplot(ss,method="pie")

```

### PerformanceAnalytics package

```{r}
library(PerformanceAnalytics)
chart.Correlation(sal1, histogram=TRUE, pch=19)

# chart.Correlation(df1, histogram=TRUE, pch=19)
```

### ggcorrplot package

```{r}
library(ggcorrplot)
ggcorrplot(cor(na.omit(sal1)), p.mat = cor_pmat(na.omit(sal1)), hc.order=TRUE, type='lower')
cor_pmat(na.omit(sal1))

ggcorrplot(cor(na.omit(df1)), p.mat = cor_pmat(na.omit(df1)), hc.order=TRUE, type='lower')
cor_pmat(na.omit(df1))
```

## ALL INCLUSIVE CORRELATION

### DataExplorer package

```{r}
library(DataExplorer)
plot_correlation(sal)

plot_correlation(df)
```

### psych package

```{r}
library(psych)
pairs.panels(sal,bg=c("yellow","blue")[sal$sex],pch=21, cex.cor=6, scale=TRUE)

# pairs.panels(df,bg=c("yellow","blue")[df$collgrad],pch=21, cex.cor=6, scale=TRUE)
```

### GGally package

```{r}
library(GGally)
ggpairs(sal)

# ggpairs(df)
```

### Test of Independence

```{r}
table(sal$rank)
table(sal$rank,sal$sex)

table(df$south)
table(df$south,df$union)

library(gmodels)
t2<-CrossTable(sal$rank,sal$sex,chisq=TRUE,fisher=TRUE)
t2

a2<-CrossTable(df$south,df$union,chisq=TRUE,fisher=TRUE)
a2
```


### T-test of Significant Diferences

```{r echo=FALSE}
filtered<-filter(sal, sex=='Male')
c1<-(filtered[6])
filtered2<-filter(sal, sex=='Female')
c2<-(filtered2[6])
t.test(c1,c2)

filtered<-filter(df, union==1)
c1<-(filtered[6])
filtered2<-filter(df, union==0)
c2<-(filtered2[6])
t.test(c1,c2)
```

### T-test with rank as the fixed effect

```{r echo=FALSE}
filtered<-filter(sal, sex=='Male',rank=='Prof')
c1<-(filtered[6])
filtered2<-filter(sal, sex=='Female', rank=='Prof')
c2<-(filtered2[6])
t.test(c1,c2)

```

## OUTLIERS

### Univariate Outlier Detection

### Interactive histogramM

```{r echo=FALSE}
library(plotly)
p1<-ggplot(sal,aes(sex,salary))+geom_boxplot()
p1
pp<-ggplotly(p1)
api_create(pp, filename = "boxplot-salary-sex")

```

### Package extremevalues

```{r echo=FALSE}
library(extremevalues)
L<-getOutliers(sal$salary)
outlierPlot(sal$salary,L)

```

### Package OutlierDetection

```{r eval=FALSE, include=FALSE}
# library(outliers)
#library(OutlierDetection)

#UnivariateOutlierDetection(sal$salary)

```

### Multivariate Outlier Detection

### Package OutlierDetection

```{r echo=FALSE}
#library(OutlierDetection)
sal1<-select(sal,salary,yrs.service)
saldata<-na.omit(sal1)
salr <- saldata[,c("yrs.service", "salary")]
#OutlierDetection(salr)

df1<-select(df,ln_wage,collgrad)
dfdata <- df1 %>% select(c(collgrad,ln_wage)) %>% na.omit()
dfr <- dfdata[,c("collgrad", "ln_wage")]
```

### Package Chemometrics

```{r echo=FALSE}
library(chemometrics)
gg<-Moutlier(saldata)
mahal_dist<-gg$md
names(mahal_dist) <- 1:nrow(saldata)
ff<-sort(mahal_dist, decreasing = TRUE)
ff

# kk<-Moutlier(dfdata)
# mahal_dist<-kk$south
# names(mahal_dist) <- 1:nrow(dfdata)
# ff<-sort(mahal_dist, decreasing = TRUE)
# ff
```

### Package Mvoutlier

```{r echo=FALSE}
library(mvoutlier)
par("mar")
par(mar=c(1,1,1,1))
out <- chisq.plot(saldata)
out$outliers

```


# AUTOMATIC

## DataExplorer

```{r echo=FALSE, include=FALSE}
library(DataExplorer)
create_report(sal,y="salary")

create_report(df,y="ln_wage")
```

## AEDA

```{r echo=FALSE,include=FALSE}
#library(efficient)
library(AEDA)
fastReport(data=sal)
rmarkdown::render("MainReport.rmd")

fastReport(data=df)
rmarkdown::render("MainReport_nlswork.rmd")
```

## ExPanDaR

```{r eval=FALSE, include=FALSE}
library(ExPanDaR)
ExPanD(sal)

ExPanD(df)
```

## RtutoR

```{r eval=FALSE, include=FALSE}
library(RtutoR)
gg=generate_exploratory_analysis_ppt(sal,target_var="salary",output_file_name="Rtutorreport.pptx")

```

