---
title: 'Data Analysis School: FORECASTING METHODS AND APPLICATIONS WITH R'
author: "Cristina Amado / UMinho"
output:
  html_document:
    df_print: paged
---

## Plan of the course

1. Introduction to Forecasting
2. Regression Analysis and Forecasting
3. Exponential Smoothing Methods
4. Univariate Time Series Models
5. Trend and Seasonality
6. Forecasting Volatility

### Setup

```{r setup}
knitr::opts_chunk$set(include = TRUE)
# install.packages('pacman')  # Package Management Tool
rm(list=ls())
library(here)
library(pacman)
start_time <- Sys.time()  # calcular o tempo de execução do programa
# setwd("/Users/miguelportela/Documents/GitHub/R_Training/time_series")
# setwd('D:\\miguel\\Dropbox\\1.miguel\\1.formacao\\r\\forecasting\\2.Lecture_Code\\_2021')
# setwd('D:\\miguel\\Dropbox\\Modelação e Análise de Dados com R - Módulo 8\\Code')
setwd('C:\\Users\\mangelo.EEG\\Documents\\GitHub\\R_Training\\timeseries\\Code')
getwd()

pacman::p_load(here,
        tidyverse,
        haven,
        #xlsx,
        #readxl,
        ggplot2,
        TSA,
        tseries,
        fma,
        forecast,
        gdata,
        lmtest,
        foreign,
        pacman,
        fpp,
        urca)
```

### Lecture 1.a

```{r Data_BEER}
# data: gold, tempdub, sales, google, oil.price, wages, larain
# data("beer",package="fma")
plot(beer,type = "l")
```


### Lecture 1.b

```{r DATA__WAGES}
data(wages)

plot(wages,type = "l")
mydata <- read.xls("sunspot.xlsx", 1)
# print(mydata)
#mydata2 = read.xls ("sunspot.xlsx", sheet = 1, header = TRUE)
```

```{r Lecture 1.b__Plot}
sunspot <- as.ts(mydata$Sunspot)
postscript(file="sunspot.eps", paper="special", width=10, height=7, horizontal=FALSE) 
plot(sunspot ~ Year, mydata, type = "l")
```

### Lecture 2.a

```{r Monthly Australian beer production: Jan 1991 - Aug 1995 (2.a)}
# data("beer", package = "fma")
plot(beer,type = "l")
```

> 80% and 95% forecast interval: level=c(80,95)

```{r Lecture 2.b__PlotBeer}
beerfor1 <- meanf(beer, h=12)
plot(beerfor1, main="mean")
beerfor2 <- naive(beer, h=12)
plot(beerfor2, main="naive")
beerfor3 <- snaive(beer, h=12)
plot(beerfor3, main="snaive")
beerfor4 <- rwf(beer, drift=TRUE, h=12)
plot(beerfor4, main="drift")
```

```{r Lecture 2.b__Accuracy}
#plot(forecast(beerfor4,level=c(80,95)))
summary(beerfor1)
summary(beerfor2)
accuracy(beerfor1)
accuracy(beerfor2)
accuracy(beerfor3)
accuracy(beerfor4)
```

### Lecture 2.b

```{r Dow-Jones index on 292 trading days ending 26 Aug 1994 (2.b)}
# data("dj",package = "fma")
plot(dj)
```

> Use 250 observations in the in-sample period

```{r Lecture 2.b__Plot}
dj1 <- window(dj, end=250)
plot(dj1)
```

#Out-of-sample period from 251 until 292
```{r Out-of-sample}
dj2 <- window(dj, start=251)
```

```{r Lecture 2.b__djfor}
#80% and 95% forecast interval: level=c(80,95)
djfor1 <- meanf(dj1, h=42)
plot(djfor1, main="mean")
djfor2 <- naive(dj1, h=42)
plot(djfor2, main="naive")
djfor3 <- snaive(dj1, h=42)
plot(djfor3, main="snaive")
djfor4 <- rwf(dj1, drift=TRUE, h=42)
plot(djfor4, main="drift")
```

> measures of forecast accuracy

```{r Measures of forecast accuracy}
accuracy(meanf(dj1,h=42), dj2)
accuracy(naive(dj1,h=42), dj2)
accuracy(snaive(dj1,h=42), dj2)
accuracy(rwf(dj1,drift=TRUE,h=42), dj2)
```

### Lecture 2.c

```{r DATA__HPRICE1}
# download.file('http://fmwww.bc.edu/ec-p/data/wooldridge/hprice1.dta','hprice1.dta',mode='wb')
hprice <- read.dta('hprice1.dta')
```

> Some statistics

```{r Some statistics}
  summary(hprice)
  cor(hprice)
  cor(hprice$price,hprice$sqrft)
```

> OLS estimation single regression

```{r Lecture 2.c__OLS estimation single regression}
lm.r <- lm(price ~ sqrft, data = hprice)
summary(lm.r)
```

> Residuals

```{r Lecture 2.c__Residuals}
resid(lm.r) #lm.r$res 
```

> Predicted values

```{r Lecture 2.c__Predicted values}
fitted(lm.r) #lm.r$fit 
```

> Scatterplot with fitted line

```{r Lecture 2.c__Scatterplot with fitted line}
plot(price ~ sqrft, hprice)
abline(lm.r) 
plot(lm.r$res ~ sqrft, hprice)
```

> Regression with logs between lprice and lsqrft

```{r Lecture 2.c__Regression with logs between lprice and lsqrft}
lm.r4 <- lm(lprice ~ lsqrft, data = hprice)
summary(lm.r4)
plot(lm.r4$res ~ lsqrft, hprice)
plot(lprice ~ lsqrft, hprice)
abline(lm.r4)
```

> Regression with logs between lprice and lassess

```{r Lecture 2.c__Regression with logs between lprice and lassess}
lm.r5 <- lm(lprice ~ lassess, data = hprice)
summary(lm.r5)
plot(lm.r5$res ~ lassess, hprice)
plot(lprice ~ lassess, hprice)
abline(lm.r5)
```

> Confidence intervals for the coefficients

```{r Lecture 2.c__Confidence intervals for the coefficients}
confint(lm.r5,level=0.95)
```

> Prediction interval for a single observation

```{r Lecture 2.c__Prediction interval for a single observation}
plot(lprice ~ lassess, hprice)
abline(lm.r5)
newdata = data.frame(lassess=250)
```

> Interval estimate on the mean at a specific point

```{r Lecture 2.c__Interval estimate on the mean at a specific point}
con.1 <- predict(lm.r5, newdata, level = 0.90, interval="confidence")
# lines(newdata,con.1[,c("lwr","upr")],col="red",lty=1,type="l")
newdata = data.frame(lassess=250)
```

> Interval estimate on a single future observation

```{r Lecture 2.c__Interval estimate on a single future observation}
pred.1 <- predict(lm.r5, newdata, level = 0.90, interval="predict")
# lines(newdata,pred.1[,c("lwr","upr")],col="blue",lty=1,type="l")
```

> Plot prediction interval

```{r Lecture 2.c__Plot prediction interval}
conf.2 <- predict(lm.r5, level = 0.95, interval="confidence")
pred.2 <- predict(lm.r5, level = 0.95, interval="predict")
fitted = pred.2[,1]
pred.lower = pred.2[,2]
pred.upper = pred.2[,3]
plot(lprice ~ lassess, hprice)
lines(hprice$lassess,fitted,col="red",lwd=2)
lines(hprice$lassess,pred.lower,lwd=1,col="blue")
lines(hprice$lassess,pred.upper,lwd=1,col="blue")
```

> OLS estimation multiple regression

```{r Lecture 2.c__OLS estimation multiple regression}
lm.r2 <- lm(price ~ sqrft + bdrms, data = hprice)
summary(lm.r2)
```

> OLS estimation multiple regression

```{r Lecture 2.c__OLS estimation multiple regression --}
lm.r3 <- lm(lprice ~ lsqrft + llotsize, data = hprice)
summary(lm.r3)
```

### Lecture 2.d

```{r Lecture 2.d__DATA}
download.file('http://fmwww.bc.edu/ec-p/data/wooldridge/wage1.dta','wage1.dta',mode='wb')
wage <- read.dta('wage1.dta')
```

```{r Lecture 2.d__Fit__Plot}
fit.r <- lm(lwage ~ educ + exper, data = wage)
summary(fit.r)
plot(fit.r$res ~ educ, wage)
plot(lwage ~ educ, wage)
abline(fit.r)
```

> Coefficients

```{r Lecture 2.d__Coefficients}
fit.r$coef #(or coef(fit.r))
```

> Residuals

```{r Lecture 2.d__Residuals}
fit.r$res 
```

> Predicted values

```{r Lecture 2.d__Predicted values}
fit.r$fit 
```

> Prediction interval for a single observation

```{r Prediction interval for a single observation}
newdata = data.frame(educ=12, exper=5)
```

> Interval estimate on the mean at a specific point

```{r Interval estimate on the mean at a specific point}
predict(fit.r, newdata, level = 0.95, interval="confidence")
```

> Interval estimate on a single future observation

```{r Interval estimate on a single future observation}
predict(fit.r, newdata, level = 0.95, interval="predict")
```

> Prediction interval for several observations

```{r Prediction interval for several observations}
neweduc=c(12,15,16)
newexper=c(5,7,10)
predict(fit.r,data.frame(educ = neweduc,exper=newexper), 
        level = 0.95, interval = "predict")
```

### Lecture 3.0

```{r Lecture 3.0__DATA}
data(elecsales, package = "fpp")
plot(elecsales,ylab="Electricity", xlab="Year")
ma.1 <- ma(elecsales,order=5)
plot(elecsales, main="Electricity",
     ylab="GWh", xlab="Year")
lines(ma(elecsales,3),col="red")
lines(ma(elecsales,9),col="blue")
print(ma.1)
print(elecsales)
```

### Lecture 3.a

```{r Lecture 3.a__DATA}
data(oil, package = "fpp")
plot(oil)
oildata <- window(oil,start=1996,end=2007)
plot(oildata, ylab="Oil (millions of tonnes)", xlab="Year")
```

> Simple exponential smoothing

```{r Simple exponential smoothing}
# lo=y1 or initial=optimal

fit1 <- ses(oildata, alpha=0.2, initial="simple", h=3)

fit2 <- ses(oildata, alpha=0.6, initial="simple", h=3)

# fit3 <- ses(oildata, h=3)
```

> Plot for alpha=0.2,0.6,0.89

```{r Plot for alpha}
plot(fit1, plot.conf=FALSE, ylab="Oil (millions of tonnes)",
     xlab="Year", main="", fcol="white", type="o")
lines(fitted(fit1), col="blue", type="o")
lines(fitted(fit2), col="red", type="o")
# lines(fitted(fit3), col="green", type="o")
lines(fit1$mean, col="blue", type="o")
lines(fit2$mean, col="red", type="o")
# lines(fit3$mean, col="green", type="o")
legend("topleft",lty=1, col=c(1,"blue","red","green"), 
       c("data", expression(alpha == 0.2), expression(alpha == 0.6),
         expression(alpha == 0.89)),pch=1)
oildata
```

> Level lt for each alpha

```{r Level lt for each alpha}
fit1$model$state
fit2$model$state

# fit3$model$state
# fitted(fit1)
# fitted(fit2)
# fitted(fit3)
```

> Summary of results

```{r Lecture 3.a__Summary of results}
summary(fit1)
summary(fit2)
# summary(fit3)
```

### Lecture 3.b

```{r Lecture 3.b__Plot}
data(ausair, package = "fpp")
#original data - from 1970 until 2010
plot(ausair) 
```

```{r Lecture 3.b__Fit}
air <- window(ausair,start=1990,end=2004)
fit1 <- holt(air, alpha=0.8, beta=0.2, initial="simple", h=5) 
fit2 <- holt(air, alpha=0.8, beta=0.2, initial="simple", exponential=TRUE, h=5) 
# fit3 <- holt(air, alpha=0.8, beta=0.2, damped=TRUE, initial="simple", h=5) 
plot(air,ylab="Air passengers", xlab="Year")
air
#print(air)
```

> Results for first model

```{r  Lecture 3.b__Results for first model}
#level lt and trend bt
fit1$model$state 
#Fitted values (lt+bt)
fitted(fit1)
```

> Forecasts

```{r Lecture 3.b__Forecasts}
fit1$mean
```

> Plots

```{r Lecture 3.b__Plots}
plot(fit2, type="o", ylab="Air passengers", xlab="Year", 
     fcol="white", plot.conf=FALSE, main="")
lines(fitted(fit1), col="blue") 
lines(fitted(fit2), col="red")
# lines(fitted(fit3), col="green")
lines(fit1$mean, col="blue", type="o") 
lines(fit2$mean, col="red", type="o")
# lines(fit3$mean, col="green", type="o")
legend("topleft", lty=1, col=c("black","blue","red","green"), 
       c("Data","Holt's linear trend","Exponential trend","Additive damped trend"))
```


> Forecast results

```{r Lecture 3.b__Forecast results}
summary(fit1)
summary(fit2)
# summary(fit3)
```

### Lecture 3.c

```{r Lecture 3.c__DATA__Plot}
data(austourists, package = "fpp")
#original data - from 1970 until 2010
plot(austourists, ylab="Tourists", xlab="Year") 
aust <- window(austourists,start=2005)
#print(aust)
```

```{r Lecture 3.c__Fit, eval=FALSE, include=FALSE}
fit1 <- hw(aust,seasonal="additive")
fit2 <- hw(aust,seasonal="multiplicative")
fit3 <- hw(aust,seasonal="additive", damped=TRUE)
fit4 <- hw(aust,seasonal="multiplicative", damped=TRUE)
#level lt, trend bt, and seasons 
fit1$model$state 
```

> Fitted values (lt+bt)

```{r Lecture 3.c__Fitted values (lt+bt), eval=FALSE, include=FALSE}
fitted(fit1)
```

> Forecasts

```{r Lecture 3.c__Forecasts, eval=FALSE, include=FALSE}
fit1$mean
plot(fit2,ylab="Tourists (millions)",
     plot.conf=FALSE, type="o", fcol="white", xlab="Year",main="")
lines(fitted(fit1), col="red", lty=2)
lines(fitted(fit2), col="green", lty=2)
lines(fit1$mean, type="o", col="red")
lines(fit2$mean, type="o", col="green")
lines(fitted(fit3), col="blue", lty=2)
lines(fitted(fit4), col="grey", lty=2)
lines(fit3$mean, type="o", col="blue")
lines(fit4$mean, type="o", col="grey")
legend("topleft",lty=1, pch=1, col=1:3, 
       c("data","Holt Winters' Additive","Holt Winters' Multiplicative"))
```

> Forecast results

```{r Lecture 3.c__Forecast results, eval=FALSE, include=FALSE}
summary(fit1)
summary(fit2)
summary(fit3)
summary(fit4)
```

### Lecture 3.d

```{r Automatic forecasting from Hyndman et al. (IJF, 2002)}
data(ausbeer, package = "fpp")
plot(ausbeer, ylab="Beer", xlab="Year") 
```

> `ets`

```{r eval=FALSE, include=FALSE}
#ets(error,trend,seasonal)
#error=A,M
#trend=N,A,Ad,M,Md
#seasonal=N,A,M
fit1 <- ets(ausbeer)
fcast1 <- forecast(fit1,h=20)
summary(fit1)
plot(fit1)
fit2 <- ets(ausbeer,model="AAA",damped=FALSE)
fcast2 <- forecast(fit2,h=20)
summary(fit2)
plot(fit2)
#plot(forecast(fit1,level=c(95),fan=TRUE))
```

### Lecture 4.0

```{r Dow-Jones index on 292 trading days ending 26 Aug 1994 ----}
data("dj", package = "fma")
```
```{r Lecture 4.0__Plot}
par(mfrow=c(1,1))
plot(dj)
retdj<-diff(log(dj))
plot(retdj)
acf(dj, lag.max = 30, na=na.omit, main="", ylab="ACF", xlab="Lag")
```

> `acf`

```{r Lecture 4.0__acf}
#acf(retdj, lag.max = 30, na=na.omit, main="", ylab="ACF", xlab="Lag",ylim=c(-0.2,0.2))
acf(retdj, lag.max = 30, na=na.omit, main="", ylab="ACF", xlab="Lag")
ar2<-arima.sim(model=list(ar=c(.9,-.2)),n=100) + 0.3
ar1<-arima.sim(model=list(ar=c(.8)),n=100) 
plot(ar2)
plot(ar1)
acf(ar1)
pacf(ar1)
acf(ar2)
pacf(ar2)
```

```{r Lecture 4.0__pacf}
ma2<-arima.sim(model=list(ma=c(-.7,.1)),n=100)
ma1<-arima.sim(model=list(ma=c(.4)),n=100) + 0.1
plot(ma1)
plot(ma2)
acf(ma1, lag.max = 40)
pacf(ma1,lag.max = 40)
acf(ma2)
pacf(ma2)
set.seed(123)
wn = rnorm(500, mean=0, sd=1)
plot(wn, type="l", ylab="wn", xlab="time")
acf(wn)
rw <- cumsum(rnorm(500, mean = 0))
plot(rw, type="l", ylab="rw", xlab="time")
acf(rw)
```

### Lecture 4.a

```{r Dow Jones Index}
# data("dj", package = "fma")
```

> Check (non)stationarity

```{r Lecture 4.a__acf}
plot(dj)
acf(dj)
pacf(dj)
```

> Augmented Dickey-Fuller

> Augmented Dickey Fuller test

```{r Lecture 4.a__adf}
adf=ur.df(dj,type="none",lags=1)
summary(adf)
```

> adf.test(dj)

```{r Lecture 4.a__adf test(dj)}
retdj <- diff(log(dj))
plot(retdj)
acf(retdj)
pacf(retdj)
```

> ARIMA(0,1,0)

```{r Lecture 4.a__ARIMA(0,1,0)}
dj.ar1 <- arima(dj, c(0, 1, 0))
summary(dj.ar1) #
acf(residuals(dj.ar1))
Box.test(residuals(dj.ar1), lag=20, fitdf=0, type="Ljung")
tsdiag(dj.ar1) 
# plot(forecast(dj.ar1)) #
```

> AR(1)

```{r AR(1)}
dj_ar1 <- arima(dj, c(1, 0, 0))
summary(dj_ar1)
acf(residuals(dj_ar1))
Box.test(residuals(dj_ar1), lag=20, fitdf=1, type="Ljung")
tsdiag(dj_ar1) 
# plot(forecast(dj_ar1))
pred.2 <- predict(dj_ar1, level = 0.95, interval="predict")
pred.2
```


### Lecture 4.b

```{r US Gross Domestic Product}
load("usgdp5190.rda")
dlgdp <- usgdp5190
```

> Check (non)stationarity

```{r Check (non)stationarity}
plot(dlgdp)
acf(dlgdp)
pacf(dlgdp)
Box.test(dlgdp, 10)
Box.test(dlgdp, 10, type = "Ljung-Box")
ar(dlgdp)
gdp.ar1 <- arima(dlgdp, c(3, 0, 0))
summary(gdp.ar1)
gdp.ar2 <- arima(dlgdp, c(1, 0, 0)) #BIC
summary(gdp.ar2)
tsdiag(gdp.ar2) 
```

> Diagnostics

```{r Diagnostics}
tsdiag(gdp.ar1) 
```

> Automatic modelling: AR models

```{r Lecture 4.b__Automatic modelling -- AR models}
gdpAR <- auto.arima(dlgdp, d=0, D=0, max.p=12, max.q=0, max.P=0, max.Q=0, 
                    max.order=12, ic="bic", stepwise=FALSE, approximation=FALSE)
summary(gdpAR)
```

> Automatic modelling: ARMA models

```{r Lecture 4.b__Automatic modelling: ARMA models}
gdpARMA <- auto.arima(dlgdp, d=0, D=0, max.p=12, max.q=12, max.P=0, max.Q=0, 
                    max.order=6, ic="aic", stepwise=FALSE, approximation=FALSE)
```

> Check with BIC

```{r Lecture 4.b__Check with BIC}
summary(gdpARMA)
tsdiag(gdpARMA)
gdp.ma2 <- arima(dlgdp, c(0, 0, 2))
summary(gdp.ma2)
```

> Diagnostics

```{r Lecture 4.b__Diagnostics}
tsdiag(gdp.ma2) 
```


### Lecture 4.c

```{r Lecture 4.c__DATA}
load("tb3m5605.rda")
tbill <- tb3m5605
```

> Check (non)stationarity

```{r Lecture 4.c__Check non-stationarity}
plot(tbill)
acf(tbill)
dltb <- diff(log(tbill))
plot(dltb)
acf(dltb)
acf(dltb^2)
pacf(dltb)
```

> Which ARMA model to choose?

```{r Lecture 4.c__Which ARMA model to choose}
tsdisplay(dltb) # forecast library
```

> Use BIC (try also a pure AR)

```{r Lecture 4.c__Use BIC (try also a pure AR)}
tbBIC <- auto.arima(dltb, d=0, D=0, max.p=10, max.q=10, max.P=0, max.Q=0,
                     max.order=10, ic="bic", stepwise=FALSE, approximation=FALSE)
summary(tbBIC)
```

> Diagnostics

```{r Lecture 4.c__Diagnostics}
tsdiag(tbBIC)
```

> Use AIC

```{r Lecture 4.c__Use AIC}
tbAIC1 <- auto.arima(dltb, max.order=5, ic="aic", stepwise=FALSE, approximation=FALSE)
summary(tbAIC1)
```

> Diagnostics

```{r Lecture 4.c__Diagnostics -}
tsdiag(tbAIC1)
```

> Try tbAIC also with max.order=12

```{r Lecture 4.c__Try tbAIC also with max.order}
tbAIC2 <- auto.arima(dltb, d=0, D=0, max.p=0, max.q=12, max.P=0, max.Q=0,
                    max.order=12, ic="aic", stepwise=FALSE, approximation=FALSE)
summary(tbAIC2)
```

> Diagnostics

```{r Lecture 4.c__Diagnostics --}
tsdiag(tbAIC2)
acf(residuals(tbAIC2))
Box.test(residuals(tbAIC2), lag=20, fitdf=7, type="Ljung")
```

> Exclude non-statistically significant coefficients

```{r Lecture 4.c__Exclude non-statistically significant coefficients}
#NA means that the coefficient is not fixed and is estimated; last coefficient 
#refers to the constant
#fl<-structure(list(mean=K,x=Dem2,fitted=tbMA0$fitted),class="forecast")
tbMA0 <- arima(dltb, order = c(0,0,7),
               fixed = c(NA, NA, NA, 0, 0, NA, NA, 0))

# summary(tbMA)
# tsdiag(tbMA)

summary(tbMA0)
tsdiag(tbMA0)

dltb_fc <- forecast(tbAIC2, h = 30)
dltb_fc
```


```{r Lecture 4.c__Plot}
#plot(gdp_fc, shaded = FALSE)
#plot(gdp_fc, shadecols = grey(c(0.8, 0.6)))
plot(dltb_fc)
```


### Lecture 4.d

> Data

```{r US Gross Domestic Product (percent change)}
load("usgdp5190.rda")
dlgdp <- usgdp5190
```

> Check (non)stationarity

```{r Lecture 4.d__Check (non)stationarity}
plot(dlgdp)
acf(dlgdp)
pacf(dlgdp)
gdp <- 100*dlgdp
plot(gdp)
gdp.ar1 <- arima(gdp, c(1, 0, 0))
summary(gdp.ar1)
```

> Predict

```{r Lecture 4.d__Predict, eval=FALSE, include=FALSE}
#pred1 <- predict(gdp.ar1, n.ahead = 4)
#pred1
gdp_fc <- forecast(gdp.ar1, h = 20)
gdp_fc
```

> Plot

```{r Lecture 4.d__Plot, eval=FALSE, include=FALSE}
#plot(gdp_fc, shaded = FALSE)
plot(gdp_fc)
```


## Close the log file

```{r End Time}
end_time <- Sys.time()
end_time - start_time
  # sprintf(end_time - start_time,fmt = '%#.1f')
```

```{r -- End Chunk --}
#
```
