---
title: "Creating reproducible research outputs"
author: "by Julia Schulte-Cloos"
date: '`r gsub("^0", "", format(Sys.time(), "%d %B, %Y"))`'
# output: beamer_presentation
output:
  # html_book
  bookdown::pdf_document2
bibliography: references.bib
biblio-style: kluwer
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r packages, echo=FALSE,results="omit",warning=FALSE,message=FALSE}
rm(list = ls())
setwd("C:/Users/mangelo.EEG/Documents/GitHub/R_Training/reproducible_research")

# Libraries

# install.packages("pacman")

pacman::p_load(
  bookdown,
  ggplot2,
  knitr,
  devtools,
  tidyverse,
  lubridate,
  modelsummary,
  showtext,
  kableExtra,
  colorspace,
  gridExtra,
  patchwork,
  xfun
)

library(haven)
library(stargazer)
```

```{r data,echo=FALSE,results="omit"}
nls <- as.data.frame(read_dta("nlswork.dta"))
```

**Abstract**

> How to succeed in making your research workflows and outputs fully reproducible? This workshop offers a gentle introduction to RMarkdown, which can be used to generate dynamic reports, write scientific articles or books, create websites, develop applications, or make interactive presentations. Participants will develop an understanding of the basic principles that underlie the framework of document generation with RMarkdown and they will be able to rely on out-of-the-box templates for their own future reproducible research outputs. 

**Agenda**

- 1. Intro - Why, when and how? The benefits of RMarkdown for research workflows

- 2. Static reproducible Rmd-documents based on LaTeX

- 3. Dynamic reproducible Rmd-documents based on CSS, Hugo, remark.js and more

- 4. Outlook: parameterized Rmd-documents)

## Slide with R Output

```{r cars}
nls %>% 
  dplyr::select(collgrad,hours) %>% 
  stargazer(title = "Summary statistics",
          label = "tb:statistcis",
          table.placement = "ht",
          header=FALSE,type="text")
```

For additional insights see @MacFarlane.

## Slide with Plot

```{r pressure}
plot(pressure)
```

# References {-}


<div id="refs"></div>

