## Factor Analysis Summary Report
Factor analysis is a way to describe variability among observed, correlated variables
in terms of a potentially lower number of unobserved variables, called **factors**. <br>
Like principal component as well as multidimensioanl scaling its purpose is dimension reduction.
It is a way to find hidden patterns, show how those patterns overlap and show what characteristics are
seen in multiple patterns.
These **factors** each embody a set of **observed variables** that have similar response patterns.
```{r loadFAObj_fa.result, echo = FALSE, message = FALSE, warning = FALSE}
if (!require("DT",character.only = TRUE)) {
    install.packages("DT",dep=TRUE)
    if(!require("DT",character.only = TRUE)) stop("Package not found")
    }
if (!require("psych",character.only = TRUE)) {
    install.packages("psych",dep=TRUE)
    if(!require("psych",character.only = TRUE)) stop("Package not found")
    }
fa.result = readRDS("fa.result.rds")
```
### Factor Analysis Summary Results

#### Pre-Analysis for optimal number of factors
In the following a pre analysis for the optimal number of factors will be displayed:

```{r showParallelAnalysis_fa.result, echo = FALSE, message = FALSE, warning = FALSE, out.width = '90%'}
plot(fa.result$task$num.fact.analysis)
```

According to the parallel analysis `r fa.result$task$num.fact.analysis$nfact ` factors should be taken.

#### Factor Loadings
The following data frame shows the factor loadings for each numeric column.
      Note that the rows are the numeric columns.
```{r showFAloadings_fa.result, echo = FALSE, message = FALSE, warning = FALSE}
loadingsWithCom = round(cbind.data.frame(as.data.frame(unclass(fa.result$fa.result$loadings)), communalities = fa.result$fa.result$communalities), 4)
dt = DT::datatable(loadingsWithCom, class = 'compact', filter = 'bottom',
      options = list(pageLength = 10), caption = htmltools::tags$caption(
      style = 'caption-side: bottom; text-align: center;',
      'Table : ', 'Factor Loadings'))
dt
```

#### Factor Analysis Graph
The following graphs visualizes the factor loadings. A cutoff of 0.3 will be applied:

```{r showFAgraph_fa.result, echo = FALSE, message = FALSE, warning = FALSE}
fa.diagram(fa.result$fa.result, cut = 0.3)
```

